---
title: "Simulation study for density and capture probability estimation from electrofishing data"
author: "Colin Millar"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r init, echo=FALSE}
knitr::opts_chunk$set(
  cache.path = "cache/simulation/",
  fig.path = "figure/simulation/",
  tidy = FALSE,
  comment = "",
  message = FALSE,
  cache = TRUE,
  eval = TRUE)
```

## Simulation code

Lets begin with a simple simulation.  First lets setup a basic data.frame from which to work.  This requires the CLdata and lubridate packages to have been installed.

```{r dataframe1}
wk <- subset(CLdata::ef, Dataset == "fobs")

# restructure
wk <- do.call(rbind, 
        lapply(c("S0", "SP", "T0", "TP"), 
            function(x) {
              out <- wk[c("Site_OBJECTID", "Site.Name", "Dataset", "Date", "Runs", "Area", paste0(x, "_R", 1:6))]
              names(out) <- gsub(x, "n", names(out))
              out $ Species <- if (substring(x, 1, 1) == "S") "Salmon" else "Trout"
              out $ LifeStage <- if (substring(x, 2, 2) == "0") "Fry" else "Parr"
              out
            }
        ))

# drop rows with all NAs
wk <- wk[apply(wk[paste0("n_R", 1:6)], 1, function(x) !all(is.na(x))), ]

# try using lubridate
wk $ Date <- as.POSIXlt(wk $ Date, tz = "GMT", format = "%d/%m/%Y")
wk $ year <- lubridate::year(wk $ Date)
wk $ doy <- lubridate::yday(wk $ Date)

# add in data summaries
wk $ Tobs <- rowSums(wk[c("n_R1", "n_R2", "n_R3")])
wk $ Xobs <- with(wk, 2*n_R1 + n_R2)
wk $ Zobs <- wk $ Xobs / wk $ Tobs

# trim data
wk <- subset(wk, doy > 150 & doy < 325 & year >= 1997 & year <= 2013 & Runs == 3 & Tobs > 0)

head(wk)
```

This provides `r nrow(wk)` observations to play with.  Okay lets simulate some 3 pass data using a constant capture probabilty, but try and get similar values for the total number of fish observed.

```{r sim1_setup, dependson = "dataframe1"}
# choose model parameter to give p of 0.6
alpha <- log(0.6/0.4)
X <- model.matrix( ~ 1, data = wk)
wk $ p <- (1 + exp(X %*% alpha))^-1
wk $ N <- round(wk $ Tobs / (1 - (1-p)^3))
n1 <- rbinom(nrow(wk), wk $ N, p)
n2 <- rbinom(nrow(wk), wk $ N - n1, p)
n3 <- rbinom(nrow(wk), wk $ N - n1 - n2, p)

wk $ T <- n1 + n2 + n3
wk $ X <- 2 * n1 + n2
wk $ Z <- wk $ X / wk $ T
```

now try to fit the model and see what we get

```{r sim1_fit}
m0 <- efp(X ~ 1, data = wk, passes = "Runs")
```

